---
title: "Migration Pattern Analysis with Data Integration"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

Import, clean and merge the datasets.

```{r}
files = list.files(
  "migration",
  pattern = "\\.xls$",
  full.names = TRUE,
  ignore.case = TRUE
)

read_migration_file = function(file_path) {
  read_excel(
    file_path,
    skip = 8,
    col_names = c(
      "dest_state_fips",
      "dest_county_fips",
      "orig_state_fips",
      "orig_county_fips",
      "state_abbr",
      "flow_description",
      "returns",
      "exemptions",
      "income_thousands"
    ),
    col_types = rep("text", 9)
  ) |>
    mutate(
      returns = parse_number(returns),
      exemptions = parse_number(exemptions),
      income_thousands = parse_number(income_thousands)
    )
}

migration_raw = map_dfr(files, read_migration_file) |> 
  arrange(dest_state_fips, dest_county_fips, orig_state_fips, orig_county_fips)
```

## 1

Identify the county pairs with the most interactions (by number of exemptions).

### Method

Before analyzing migration patterns, I removed rows that did not represent meaningful county-to-county migration. Specifically, I excluded summary rows where orig_state_fips > 56, since these rows correspond to aggregated categories rather than actual U.S. counties. I also removed rows where the origin and destination counties were the same, because this project focuses on migration between counties rather than within-county records.

To identify the county pairs with the most interactions, I first created standardized county identifiers by concatenating state and county FIPS codes for both origin and destination counties. Because county-to-county migration is directional in the raw data, I reordered each pair using `pmin()` and `pmax()` so that flows between the same two counties would be treated as one undirected county pair. I then aggregated the number of exemptions across each pair and ranked the pairs from highest to lowest interaction.

To make the output interpretable, I also created a county lookup table from the non-migrant rows in the original dataset. I then joined this lookup table back to the ranked county pairs so that the final results included both FIPS codes and county names.

```{r}
# get rid of the summary rows and non-migration rows
migration_df = migration_raw |> 
  filter(
    orig_state_fips <= 56,
    !(dest_state_fips == orig_state_fips & dest_county_fips == orig_county_fips)
  ) |> 
  mutate(
    # create unique FIPS codes for each county
    dest_fips = paste0(dest_state_fips, dest_county_fips),
    orig_fips = paste0(orig_state_fips, orig_county_fips))

county_pairs = migration_df |>
  mutate(
    # create county pairs
    pair_1_fips = pmin(orig_fips, dest_fips),
    pair_2_fips = pmax(orig_fips, dest_fips)
  ) |>
  # sum exemptions for each county pair
  group_by(pair_1_fips, pair_2_fips) |>
  summarise(
    total_exemptions = sum(exemptions, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_exemptions))
```

```{r}
# create a county name lookup table
county_lookup = 
  migration_raw |> 
  filter(dest_state_fips == orig_state_fips & dest_county_fips == orig_county_fips) |>
  mutate(
    county_fips = paste0(dest_state_fips, dest_county_fips),
    county_name = str_remove(flow_description, " Non-Migrants$"),
    county_state = paste0(county_name, ", ", state_abbr)
  ) |>
  select(dest_state_fips, dest_county_fips, county_fips, state_abbr, county_state) |>
  distinct()

# join county names to each county pair
top_pairs_named = county_pairs |>
  left_join(
    county_lookup |>
      rename(
        pair_1_fips = county_fips,
        pair_1_state = state_abbr,
        pair_1_county = county_state
      ),
    by = "pair_1_fips"
  ) |>
  left_join(
    county_lookup |>
      rename(
        pair_2_fips = county_fips,
        pair_2_state = state_abbr,
        pair_2_county = county_state
      ),
    by = "pair_2_fips"
  ) |> 
  select(pair_1_fips, pair_1_county, pair_2_fips, pair_2_county, total_exemptions)

top_pairs_named |> 
  head(5) |> 
  knitr::kable(caption = "Table 1. Top 5 county pairs by total exemptions") |> 
  kableExtra::kable_styling(full_width = FALSE)
```


## 2

Determine which counties show the highest in-migration and out-migration.

### Method

To identify the counties with the highest in-migration and out-migration, I aggregated migration flows at the county level using the number of exemptions as the measure of migration volume. Specifically, I summed exemptions across all records for each destination county (dest_fips) to measure total in-migration, and across all records for each origin county (orig_fips) to measure total out-migration. After ranking counties in descending order of total exemptions, I joined the results with the county lookup table to attach readable county names and state abbreviations.

```{r}
# counties with the highest in-migration
top_in_migration = migration_df |>
  group_by(dest_fips) |>
  summarise(
    total_in_exemptions = sum(exemptions, na.rm = TRUE)
  ) |>
  arrange(desc(total_in_exemptions)) |> 
  left_join(county_lookup, by = c("dest_fips" = "county_fips")) |> 
  select(dest_fips, county_state, total_in_exemptions)

top_in_migration |> 
  head(5) |> 
  knitr::kable(caption = "Table 2. Counties with the highest in-migration") |> 
  kableExtra::kable_styling(full_width = FALSE)

# counties with the highest out-migration
top_out_migration = migration_df |>
  group_by(orig_fips) |>
  summarise(
    total_out_exemptions = sum(exemptions, na.rm = TRUE)
  ) |>
  arrange(desc(total_out_exemptions)) |> 
  left_join(county_lookup, by = c("orig_fips" = "county_fips")) |> 
  select(orig_fips, county_state, total_out_exemptions)

top_out_migration |> 
  head(5) |> 
  knitr::kable(caption = "Table 3. Counties with the highest out-migration") |> 
  kableExtra::kable_styling(full_width = FALSE)
```












